Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)

### 2017-05-11 WSSSPE4_report/WSSSPE4_paper.tex
more shortening removing most appendices removing talk descriptions and combining all talks into one section shortening another working group subsection

\begin{abstract}
This report records and discusses the Fourth Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE4). The report
includes a description of the keynote presentation of the workshop, the
mission and vision statements that were drafted at the workshop and
finalized shortly after it, a set of idea papers, position papers, experience
papers, demos, and lightning talks, and a panel discussion.  The main part of the report
covers the set of working groups that formed during the meeting, and for each,
discusses the participants, the objective and goal, and how the objective can
be reached, along with contact information for readers who may want to join
the group. Finally, we present results from a survey of the workshop attendees.

@@

\todo{cite~\cite{WSSSPE4-report}}

@@

WSSSPE4 also included two professional event organizers/facilitators from
KnowInnovation who helped the organizing committee plan the workshop agenda,
and during the workshop, they actively engaged participants with various
tools, activities, and reminders.

@@

This report is based on the events at the workshop and the submitted materials.
These events included discussion of the call for papers (\S\ref{sec:preworkshop}), the WSSSPE mission and vision (\S\ref{sec:mission}), a keynote (\S\ref{sec:keynote}),
a set of presentations (\S\ref{sec:papers} and \S\ref{sec:lightning}), a panel (\S\ref{sec:panel}), and a number of working groups (\S\ref{sec:WGs}).  One full day of the
workshop was spent with participants in the working groups, which occurred in parallel
with each other.  And each of the working groups left with a plan for how they could move
forward.
This report also mentions the Slack channel created for further discussions (\S\ref{sec:slack}), and it highlights an attendee survey (\S\ref{sec:survey}) before concluding (\S\ref{sec:conclusions}).
Appendices list the organizers (\S\ref{sec:orgcom}), attendees(\S\ref{sec:attendees}), travel award recipients (\S\ref{sec:awardees}), and program committee (\S\ref{sec:progcom}), and detail the attendee survey (\S\ref{sec:survey_details}).

@@

Submissions to WSSSPE4 comprised
19 lightning talks,
4 idea papers,
3 position paper,
5 experience papers,
and
3 demos.
Most of the submissions were accepted, since the goal of WSSSPE is always to
take in as many relevant inputs as possible, and to encourage their authors to
participate in sharing and implementing their ideas.
Specifically,
19 lightning talks,
4 idea papers,
3 position papers,
2 experience papers,
and
3 demos
were accepted. (Two of the submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
were accepted as lightning talks.)
The papers are discussed in Section~\ref{sec:papers},
and the lightning talks are discussed in Section~\ref{sec:lightning}.
The paper and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Keynote}\label{sec:keynote}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The keynote was given by Patricia Lago and entitled ``The legacy of unsustainable software''.
%
Sustainability is broadly associated with natural ecologic systems. When we translate the notion of sustainability to software solutions, however, we often confuse {\em impact in a certain point in time} with {\em balanced and durable effects}. In addition, software sustainability adds a fourth dimension to environmental, social and economic aspects: technical sustainability, and hence higher complexity~\cite{Lago2015}.

Supporting technical sustainability in (scientific) software is very near to supporting properties like longevity and resilience, i.e., the ability of a software system or application to fulfill its intent~\cite{Huisman2016}, as planned independently from time passing or from changes in its operational context. Technical sustainability has never been easy, but it is even more of a challenge nowadays with the increasing demands on software solutions to support many aspects of our daily lives.

The keynote discussed the challenges (and some related ongoing research) of combining technical and environmental sustainability. This provided a complementary angle to the workshop discussions. It addressed software energy efficiency (i.e., how to develop software that consumes less energy, like mobile apps that should consume less device-battery) and software energy-awareness (i.e., how to develop software that can actively decide between energy efficiency and other goals such as performance).

The keynote used a few myths to elaborate on such challenges and related research. Among them:

\begin{itemize}
\item[] {\bf Myth \#1: Energy-efficient hardware will solve the issue}. At the implementation level, to make software use resources effectively, we need to first understand the impact of different coding practices on the way software exploits computing resources (hence energy to feed them). In addition, when we have to deal with complex software or new projects, we must understand ``the big picture,'' hence we must reason at the design or architecture level. Maybe the software is not implemented yet, or maybe we must identify which design decisions are more or less beneficial for energy consumption. To this aim, we capture best practices in reusable patterns and so-called architectural tactics. These document generic designs, specific implementations, metrics, and measures.\\

\item[] {\bf Myth \#2: Extensive experimentation will provide the needed know how}. While empirical methods to design and execute software engineering experiments are mature and well known, they are insufficient to handle the too many variables and uncertainties for modern software to be sustainable. We need new empirical methods that help us gather evidence faster~\cite{Procaccianti2015}. This would probably entail relaxing the too many constraints that govern current rules for mitigating threats to validity, and (equally importantly) collaboration- and experience-sharing to build a common knowledge base. In addition, we should exploit better (and more truthful) visualizations of experimentation results, to communicate on advances (and less promising directions.)\\

\item[] {\bf Myth \#3: With massive migration of software systems to cloud-based environments, we do not need to invest in a ``greener cloud'' anymore}. The use of cloud-based services promises large-scale optimizations in terms of the resources used by our software applications and related investments~\cite{green-cloud-sw}. Unfortunately, cloud provisioning heavily relies on over-capacity rather than intelligent governance, and so far has invested very little in innovation for sustainability. We need to build smarter software that both manages resources more efficiently in the cloud, better understands the changes in the software operational context (including user requirements), and dynamically adapts to address sustainability intents~\cite{Procaccianti2016-2} (both in the cloud and on customer's premises.)\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Papers and demos} \label{sec:papers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

WSSSPE included the presentation of 12 10-minute talks (based on four idea papers, three position papers,
two experience papers, and three demos) that addressed a wide range of topics around
sustainability for software in science. These talks covered three main areas:
\begin{enumerate}
\item The specifics of the academic environment for software development and for so-called
research programmers or research software engineers;
\item The characteristics and needs of research software in general such as the lifecycle of
software in science, preservation, curation, sharing, and reproducibility challenges; and
\item Lessons learned from or visions for use cases of scientific software and communities
working with and\slash or on a software package.
\end{enumerate}

%
%
%\todo{group by type - idea, position, experience, demo}

The following four papers belong to the first area: the academic environment.
They address diverse aspects ranging from incentives for quality software to
dvocating a professional society for research software to roles and degrees for
research software engineers. For talks with multiple authors, an
\textsuperscript{\textasteriskcentered} indicates the author who presented the talk.

\begin{itemize}
\item \textbf{Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
Improving Incentives for Quality Software}}~\cite{Heroux:2016ws}.
The author presents that improving quality of scientific software is a measure
to address the increased complexity in both modeling and software. To support
quality, it is essential to provide Computational Science and Engineering teams
with publication venues, funding, and  professional recognition. Broad
improvements in sustainability and productivity can be only achieved when
publishers, funding agencies and employers raise their expectations for
software quality. Additionally, software community leaders are in the unique
position to positively impact software quality by working on establishing
incentives that will spur creative and novel approaches to improve developer
productivity  and software sustainability.

\item \textbf{Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
Society for Research Software}}~\cite{GAllen:2016ws}.
This paper advocates to create a professional society or extend the scope of an
existing organization in research software. The author presents the motivation,
a variety of potential roles of such a society and diverse potential funding
mechanisms to be able to establish a sustainable organization. She gives an
overview on similar activities as well as existing organizations and aims at
stimulating discussion among potential members and stakeholders.

\item \textbf{Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
\emph{Experience Paper: Preliminary analysis of a survey of UK Research
Software Engineers}}~\cite{Philippe:2016ws}.
This paper presents results from a survey conducted on the new role of research
software engineers in academia. The survey provides demographic information
about the education, field, gender, job satisfaction, and career plans of the people
in this community. The outcome is that members of the community  were found to be
highly educated, derive mainly from the hard sciences, and are predominantly male.
Respondents reported satisfaction in their jobs, but indicated that career progression
is both difficult and opaque. The authors urge a continued discussion about the experience
of research software engineers and recommend further investigation into this
important community.

\item \textbf{Christopher Gwilliams: \emph{Demo: Using industrial engagement to
create and develop research ties within academia}}~\cite{Gwilliams:2016ws}.
This demo focuses on the benefits of including a  new  software engineering
degree that is influenced by, and works directly with,
industry.  Cardiff  University  introduced  a  new  industry-focused  degree,
Bachelor of Science in Applied Software Engineering (ASE) in September
2015. The engagement with industry allows students to learn through live
projects  with  real-world  applications and forge links for career opportunities.
Also, industry benefits from this course that provides an entry point for industry
to engage with the university and to develop a network of contacts between industry
and academia, which is also a benefit for the university and allows to identify
potential opportunities for new research projects and collaborators. \\

The second main area, concerned with characteristics and needs of research software, includes the following five papers:

\item \textbf{Anshu Dubey\textsuperscript{\textasteriskcentered} and Katherine Riley: \emph{Experience Paper: Software
Engineering and Community Codes Track in ATPESC}}~\cite{Dubey1:2016ws}.
The Argonne Training Program in Extreme Scale
Computing (ATPESC) was started by Argonne National Laboratory with the objective
of expanding the ranks of better prepared users of high-performance computing (HPC) machines.
One of the unique aspects of the program was inclusion of a
software engineering and community codes track. The inclusion
was motivated by the observation that the projects with a good
software process were better able to meet their scientific goals.
The paper presents the experience in running the software
track, the motivation, reception, and evolution of the track over the years.
The authors recommend the inclusion of similar tracks in other HPC-oriented training programs.

\item \textbf{Neil Chue Hong: \emph{Position Paper: Why do we need to compare
research software, and how should we do it?}}~\cite{ChueHong:2016ws}.
This position paper sets out the reasons for why different stakeholders,
from users to developers to funders, might wish to undertake the difficult task
to compare research software. The author describes a proposed framework for doing
so (based around measures of accessibility, usability, maintainability, and
portability) which takes into account the possibility of variation between
different communities about how they prioritize different aspects of research software.

\item \textbf{Anshu Dubey\textsuperscript{\textasteriskcentered} and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
of Software for Scientific Simulation}}~\cite{Dubey2:2016ws}.
The authors discuss the specifics of the lifecycle of scientific software.
While the software lifecycle is a well-researched topic, one class of projects,
research software, has received relatively
little attention from lifecycle researchers. This
paper formalizes a lifecycle model
for end-to-end scientific application software, featuring loose
coupling between submodels for development of infrastructure
and scientific capability. The authors also invite input from stakeholders
to converge on a model that captures the complexity of this
development processes and provides needed lifecycle guidance to
the scientific software community.

\item \textbf{Francisco Queiroz\textsuperscript{\textasteriskcentered} and Rejane Spitz: \emph{Position Paper:
Collaborative Gamification Design for Scientific Software}}~\cite{Queiroz:2016ws}.
The paper focuses on usability design and presents that gamification, a design
trend for improving scientific software usability, for scientific software should
be facilitated by an open, collaborative design process supported by
conversational media. This approach is compatible with
qualities often attributed to computational science community
regarding openness and collaboration between members of varied
professional backgrounds. They exemplify the use of conversational media for
collaborative design and expect the synergy between collaborators to result in
better usability, greater user acceptance, and adequacy to requirements,
obtaining optimal design solutions in a sustainable way.

\item \textbf{Bruce Childers\textsuperscript{\textasteriskcentered}, Jack Davidson, Wayne Graves, Bernard Rous, and
David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
the Way Digital Libraries will Operate}}~\cite{Childers:2016ws}.
The paper presents the implications of ``active curation systems'' that provide
executable access to
artifacts and experiments behind published results and enables
their reuse. These systems allow changing and repeating experiments to understand
how an innovation behaves in conditions beyond the ones described in a paper and
address reproducibility challenges. Active curation systems
also  enable accountability and accelerate research progress  by
giving access to complete experimental details. As these systems
take  hold,  it  is  important  to  understand  their  capabilities  and
how digital libraries should be integrated with them. The authors describe a
study underway to explore how best
to do this integration.\\

Three contributions focused on the third main area: elucidating lessons learned
from or visions for use cases of scientific software and communities working
with and\slash or on a software package.

\item \textbf{Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse\textsuperscript{\textasteriskcentered}: \emph{Demo: Composing,
Reproducing, and Sharing Simulations}}~\cite{Ganguly:2016ws}.
The authors demonstrate their approach of sharing,
reproducing, and composing simulations toward accelerating
research productivity while also improving accountability and
credibility. Specifically, they have developed a case study in
which they compose and share access control simulations in the
form of shareable data store units for cloud systems. They share, compose
and repeat simulations  through a collaborative repository
(OCCAM) and suggest a general simulation framework (SST) that can
accelerate the efforts for the whole community.

\item \textbf{Sameer Shende\textsuperscript{\textasteriskcentered} and Allen Malony: \emph{Demo: Using TAU for
Performance Evaluation of Scientific Software}}~\cite{Shende:2016ws}.
This demo presents the  TAU Performance System for performance evaluation of Scientific
Software written in C++, C, and Fortran. The system addresses performance
technology problems  at three levels: instrumentation,
measurement, and analysis. The TAU framework supports the
configuration and integration of these layers. The authors advocate that effective
exploration of performance will necessarily
require prudent selection from the range of alternative methods TAU provides
to  assemble meaningful performance experiments.

\item \textbf{Janos Sallai\textsuperscript{\textasteriskcentered}, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
\emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
Atom-Typing for Molecular Simulation}}~\cite{Sallai:2016ws}.
The authors present an essential and error-prone task in the area of Molecular
Dynamics simulations: the efficient selection of correct parameter values for
specific molecules, especially forcefields. They present a framework that
aims to solve this data management issue, proposing a common format for
forcefields that is self-documenting with machine readable,
declarative usage rules. They investigate processes and tools
that are commonly used today in software development (e.g.,
unit testing, verification and validation, continuous integration,
and version control) and reason that they are, with proper infrastructure
support, applicable to forcefield development, as well. The
paper describes how such an infrastructure can tackle managing and evolving
forcefields by the molecular dynamics community, and
proposes a way to encourage and incentivize involvement
by the stakeholders.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Lightning talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}
\note{
\href{http://wssspe.researchcomputing.org.uk/wssspe4/agenda/}{Slides.}}
\end{comment}

There were 19 lightning talks presented at WSSSPE4, as follows (in order of
presentation).
%
For talks with multiple authors, an \textsuperscript{\textasteriskcentered} again indicates the author who presented the talk.
%
\begin{itemize}[itemsep=1ex]
    \item \textbf{Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
    programme; supporting the social side of research software}}~\cite{Sufi:2016ws}.
    Sufi, Community Lead at the Software Sustainability Institute (SSI), discussed
    the SSI Fellowship program and the fellows, including their
    demographic breakdown and domains of practice. This program came about
    based on the institute's desire to integrate with the research community---thus
    inspiring change from within---rather than preach to it. Five SSI fellows
    attended WSSSPE4.

    \item \textbf{Sandra Gesing\textsuperscript{\textasteriskcentered}, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
    Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
    \emph{The Science Gateways Community Institute - Supporting Communities to
    Achieve Sustainability for Their Science Gateways}}~\cite{Gesing:2016ws}.
    In this talk, Gesing introduced the concept of science gateways: end-to-end software
    solutions for a community that hide underlying complexity. She also posed the need
    for gateways, which include the increased complexity of research and associated
    hardware\slash software, a greater need for reproducibility and openness, and the
    opportunity to integrate research with teaching for better workforce development.
    In fact, gateway users surpassed login-based users in XSEDE systems in 2013,
    and 1.4 times as many users accessed XSEDE via gateways compared with
    login in March 2016. The Science Gateways Community Institute offers support
    related to gateways in five areas: Incubator, Extended Developer Support, Scientific
    Software Collaborative, Community Engagement and Exchange, and Workforce
    Development.

    \item \textbf{Karthik Ram\textsuperscript{\textasteriskcentered}, Noam Ross,
    and Scott Chamberlain: \emph{A model for
    peer review and onboarding research software}}~\cite{Ram:2016ws}.
    This talk described the rOpenSci community's review process for research
    software packages. Submission of a package starts with a pre-submission inquiry
    for fit with the community criteria, then peer reviewers evaluate based on
    usability, quality, and style. Upon acceptance, packages are given an rOpenSci
    README badge and added to the system. Important components include a code of
    conduct, open and non-adversarial review, and that no submission is
    rejected---just out of scope. Remaining challenges include scaling, author
    incentives, and automation of the process.

    \item \textbf{Frank L\"{o}ffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
    \emph{A vision of computing in 10+ years}}~\cite{Loffler:2016ws}.
    This talk began by describing the challenges facing science, and particularly
    computational science, in recent years: the amount and complexity of knowledge
    has increased dramatically, making it difficult for individuals or even teams to
    manage modern, complex problems.
    Brandt submitted an argument for decoupling science simulation and
    the solution of computational problems away from numerical methods and low-level
    optimization. This may be possible with domain-specific languages, or the use
    of modern features in existing languages. This vision also includes reuse of
    basic computing infrastructure, including common data structures, parameter files,
    and high-performance I\slash O. This computing future necessarily requires
    better recognition and reward systems for research software developers (i.e.,
    research software engineers), and ways to make increasingly complex scientific
    papers more understandable.

    \item \textbf{Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
    Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
    \emph{Software Citation: Process, Principles, and Implementation}}~\cite{Katz:2016ws}.
    This talk described the timeline of activities of the FORCE11 Software Citation
    Working Group and the principles of software citation established by the group.
    Niemeyer, one of the working group co-chairs, also discussed the paper published
    describing the principles~\cite{Smith:2016sc}, and the remaining steps of the
    working group before sunsetting it and spinning up a new Software Citation
    Implementation Working Group.

    \item \textbf{Ray Idaszak\textsuperscript{\textasteriskcentered},
    David G.~Tarboton, Hong Yi, Michael Stealey,
    Pabitra Dash, Alva Couch, Daniel P.~Ames, Jeffery S.~Horsburgh, Tony Castronova,
    Jon Goodall, Mohamed Morsy, Venkatesh Merwade, Mauriel Ramirez, Tian Gan,
    Drew (Zhiyu) Li, Jeff Sadler, Shawn Crawley, Zhaokun Xue, Lan Zhao, Carol Song,
    Christina Bandaragoda: \emph{HydroShare - A Case Study in Software Engineering
    Best Practices and Culture Change for Developing Sustainable Community
    Software}}~\cite{Idaszak:2016ws}.
    HydroShare is a five-year-old online, collaborative system that supports
    open sharing of hydrologic data, analytical tools, and computer models. The
    project involves 10 collaborating institutions that produce tested and reviewed
    community code every two to three weeks, following software engineering best
    practices. Undergraduate and graduate students involved in the project not
    only adopt the best practices exhibited by HydroShare, but see the open,
    collaborative approach to research software development as the norm.
    The talk made recommendations to other community efforts, including the need to
    identify committed, hands-on PIs; evidence-based software engineering for science;
    continuous integration; the importance of professional software development and
    operations practices, even for domain scientists; and to identify technology
    decisions that overlap with career paths.

    \item \textbf{Lorraine Hwang\textsuperscript{\textasteriskcentered},
    Wolfgang Bangerth, Timo Heister, and Louise Kellogg:
    \emph{ASPECT: Hackathons as an Example of Sustaining an Open Source
    Community}}~\cite{Hwang:2016ws}.
    In this talk, Hwang first described the ASPECT (Advance Solver for Problems in
    Earth ConvecTion) open-source, parallel, extensible finite element code
    used to simulate thermal convection. ASPECT began development in 2011 as a
    community-driven project under the Computational Infrastructure for Geodynamics
    (CIG). Community-building activities around ASPECT focused on the traditional
    workshops, tutorials, and webinars, but also on slightly nontraditional,
    multi-day hackathons---which turned out to be one of the most effective
    forms of fostering a community. These start with leaders establishing project
    conventions, demonstrating workflows and relevant technical skills, and
    identifying starter projects. Daily activities involve work plans and
    science\slash coding briefs, with mini-tutorials given as needed. One important
    requirement is that each participant must submit at least one commit. Hwang
    also summarized lessons learned, including that hackathons should last 7--10
    days, include participants with multidisciplinary expertise and at both novice
    and expert levels, have at least one dedicated code reviewer per 10 participants,
    hold daily group updates, organize activities to promote group interactions,
    and enforce free time.

    \item \textbf{Carole Goble: \emph{A Simple Profiling Framework for Software
    User-Producer Reciprocity Review}}~\cite{Goble:2016ws}.
    Goble first described some existing frameworks for measuring the maturity of
    software, including the Software Sustainability Maturity Model, OSS Watch
    Openness Rating, NASA Reuse Readiness Rating, Capability Maturity Model, and
    Qualification and Selection of Open Source software. Then, she proposed a new,
    simpler framework for project reciprocity, drawing on the intentions of users
    and producers of the software, from the viewpoint of the producer.
    Next steps include further definition of the model's characteristics based on
    both theoretical and empirical investigation.

    \item \textbf{Stephan Druskat: \emph{A proposal for the measurement and
    documentation of research software sustainability in interactive metadata
    repositories}}~\cite{Druskat:2016ws}.
    In this talk, Druskat first identified two technical barriers to software
    sustainability: identification of good software, and software discoverability.
    Some of his perspectives were based on experiences in the humanities, a lack
    of funding for computational infrastructure results in little experience with
    sustainability (although this may differ in other domains). Druskat proposed
    metadata repositories as a possible solution, as they can measure and document
    technical sustainability, and document resulting metrics and more general
    features of software. Technical sustainability goals include ensuring software
    existence, preserving potential operation of software, and creating\slash
    retaining possibilities for further development and maintenance of software.
    The next steps of such a system include compiling measurable criteria for
    the sustainability goals (based on qualitative concepts of, e.g.,
    ``sustainability,'' ``maintainability,'' and ``usability''), and
    crowd-sourcing a weighting of these criteria.

    \item \textbf{Mistral Contrastin, Matthew Danish,
    Dominic Orchard\textsuperscript{\textasteriskcentered}, and
    Andrew Rice: \emph{Supporting software sustainability with
    lightweight specifications}}~\cite{Contrastin:2016ws}.
    This talk introduced the tool CamFort for verifying Fortran code based on
    lightweight specifications, which aid in software maintainability and reuse
    through high-level indications of programmer intent.
    Currently, CamFort supports specification (and
    verification) of units and stencil operations. Plans for future specifications
    include automatic test generation based on user-supplied properties, dependency
    specifications, and software contracts to check, e.g., expected value ranges
    and program behavior.

    \item \textbf{Alice Allen, Cecilia Aragon, Christoph Becker, Jeffrey Carver,
    Andrei Chi\c{s}, Benoit Combemale, Mike Croucher, Kevin Crowston, Daniel Garijo,
    Ashish Gehani, Carole Goble\textsuperscript{\textasteriskcentered},
    Robert Haines, Robert Hirschfeld, James Howison,
    Kathryn Huff, Caroline Jay, Daniel S.~Katz, Claude Kirchner, Kateryna Kuksenok,
    Ralf L\"{a}mmel, Oscar Nierstrasz, Matt Turk, Rob van Nieuwpoort, Matthew Vaughn,
    and Jurgen Vinju: \emph{``I solemnly pledge'': A Manifesto for Personal
    Responsibility in the Engineering of Academic Software}}~\cite{AAllen:2016ws}.
    This talk described the activities of the weeklong Dagstuhl Perspectives Workshop
    manifesto on ``Engineering Academic Software,'' which produced a manifesto
    describing a pledge affirming the scholarly value of research software and
    the personal responsibility to act in accordance with that value. The draft
    (personal) manifesto includes 20 pledges, organized into three areas:
    recognition of academic software, academic software development processes,
    and the intellectual content of academic software. The talk concluded with a
    survey for participants to give feedback on the pledges and their importance.

    \item \textbf{Neil Chue Hong\textsuperscript{\textasteriskcentered}:
    \textit{Making it easier to understand research software impact}}~\cite{ChueHong:2016wsb}.
    Chue Hong proposed that the WSSSPE community collaboratively develop a framework
    to better understand (and measure the impact of) research software. This could
    be based on a workflow involving mining the literature for mentions of software,
    identifying linkages between software, and visualizing dependencies between
    software or between software and other research artifacts. Many tools already
    exist in this ecosystem, so Chue Hong proposed that a WSSSPE working group
    collect these tools, summarize use cases, identify missing pieces, and find
    low-hanging fruit to integrate tools or build new ones.

    \item \textbf{Willem Robert van Hage\textsuperscript{\textasteriskcentered},
    Jason Maassen and Rob van Nieuwpoort: \textit{Software Impact Measurement at the
    Netherlands eScience Center}}~\cite{vanHage:2016ws}.
    This talk described eStep (eScience Technology Platform) at the Netherlands
    eScience Center, which includes three components: a software catalogue
    of eScience software, interfaces, libraries,
    tools workflows, applications, etc.; a knowledge base of guides, reports, and
    recommendations for scientific software development; and eScience research
    documents including such as scientific publications, book chapters, and demos.
    The eStep website currently presents automatically-generated information about
    software based on metadata, with a plan for displaying measures of impact and
    quality.

    \item \textbf{Gabrielle Allen, Emily Chen\textsuperscript{\textasteriskcentered},
    Ray Idaszak, and Daniel S.\ Katz: \textit{Report on Software Metrics for
    Research Software}}~\cite{GAllen:2016wsb}.
    In this talk, Chen first summarized some of the outcomes of the WSSSPE3
    Metrics Breakout Group, who discussed that metrics are important for scientific
    impact, discovery, reducing duplication of efforts, and evaluating software.
    The group also summarized barriers to widespread use of metrics: no common
    set of metrics collected, a need for usefulness to be obvious for projects,
    and no common standard for collecting\slash presenting metrics. To find out
    more about practices related to metrics, the authors ran a survey of National
    Science Foundation Software Infrastructure for Sustained Innovation (SI2)
    program investigators, with 19 participants. Important results include the
    observation that most use citations and acknowledgements of software in
    funding proposals, while download numbers are both challenging to collect
    and not useful. Overall, participants generally agreed that metrics are
    important, but said useful metrics are difficult or tedious to collect.

    \item \textbf{Eric L.~Seidel\textsuperscript{\textasteriskcentered}
    and Gabrielle Allen: \textit{Bringing Techniques from Software Engineering
    into Scientific Software}}~\cite{Seidel:2016ws}.
    This talk, given by a senior graduate student in Computer Science\slash
    programming languages research, described several areas of software engineering
    concepts that benefit the development of research software. In particular,
    Seidel focused on software correctness, starting from high-level
    specifications. Correctness can be demonstrated by verifying or proving
    that the software satisfies the specification, testing against a range of
    inputs, or synthesis of code from the specification. However, all of these
    present technical challenges, in addition to the social challenges of
    encouraging collaboration between software engineering and research software
    communities.

    \item \textbf{Iain Emsley\textsuperscript{\textasteriskcentered} and David
    De Roure: \textit{Sustaining the social: Connecting the lives of Drupal
    community}}~\cite{Emsley:2016ws}.
    In this talk, Emsley discussed efforts to develop the community around the
    Drupal content management system, which is supported through the DrupalCon
    series of conferences, user groups, and an association.
    The talk described the use of social coding and software credit to develop
    the Drupal software community, and social documentation to provide
    documentation and tutorials. By analyzing identities of committers, they
    found that the larger community is actually a loosely coupled set of
    heterogeneous communities that use (and sustain themselves) through different
    social mechanisms.

>>>>

\begin{abstract}
This report records and discusses the Fourth Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE4). The report
includes a description of the keynote presentation of the workshop, the
mission and vision statements that were drafted at the workshop and
finalized shortly after it, a set of idea papers, position papers, experience
papers, demos, and lightning talks, and a panel discussion.  The main part of the report
covers the set of working groups that formed during the meeting, along with contact information for readers who may want to join
the group. Finally, we present results from a survey of the workshop attendees.

@@

@@

WSSSPE4 also included two professional event organizers/facilitators who helped 
plan the workshop agenda,
and who engaged with participants during the workshop.

@@

This paper is a summary of a longer report~\cite{WSSSPE4-report} on the events at the workshop and the submitted materials.
The remainder of the paper includes the call for papers (\S\ref{sec:preworkshop}); the WSSSPE mission and vision (\S\ref{sec:mission}); a set of presentations that included a keynote, papers, lighting talks, and a panel (\S\ref{sec:presentations}); and the activities of a set of working groups (\S\ref{sec:WGs}).  One full day of the
workshop was spent with participants in the working groups, which occurred in parallel
with each other.  And each of the working groups left with a plan for how they could move
forward.
This report also mentions the Slack channel created for further discussions (\S\ref{sec:slack}), and it highlights an attendee survey (\S\ref{sec:survey}) before concluding (\S\ref{sec:conclusions}).
An appendix lists the organizers and program committee members (\S\ref{sec:orgcom}).

@@

Submissions to WSSSPE4 comprised
19 lightning talks,
4 idea papers,
3 position paper,
5 experience papers,
and
3 demos.
Most of the submissions were accepted, since the goal of WSSSPE is always to
take in as many relevant inputs as possible, and to encourage their authors to
participate in sharing and implementing their ideas.
Specifically,
19 lightning talks,
4 idea papers,
3 position papers,
2 experience papers,
and
3 demos
were accepted, as mentioned in Section~\ref{sec:presentations}. (Two of the submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
were accepted as lightning talks.)
The paper and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Presentations}\label{sec:presentations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The keynote was given by Patricia Lago and entitled ``The legacy of unsustainable software''.
%
Sustainability is broadly associated with natural ecologic systems. When we translate the notion of sustainability to software solutions, however, we often confuse {\em impact in a certain point in time} with {\em balanced and durable effects}. In addition, software sustainability adds a fourth dimension to environmental, social and economic aspects: technical sustainability, and hence higher complexity~\cite{Lago2015}.  Lago's keynote discussed the challenges (and some related ongoing research) of combining technical and environmental sustainability, providing a complementary angle to the workshop discussions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Papers and demos} \label{sec:papers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

WSSSPE included the presentation of 12 10-minute talks (based on four idea papers, three position papers,
two experience papers, and three demos) that addressed a wide range of topics around
sustainability for software in science. These talks covered three main areas.

The following four papers belong to the first area: the academic environment.
They address diverse aspects ranging from incentives for quality software to
advocating a professional society for research software to roles and degrees for
research software engineers. For talks with multiple authors, an
\textsuperscript{\textasteriskcentered} indicates the author who presented the talk.

\begin{itemize}
\item Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
Improving Incentives for Quality Software}~\cite{Heroux:2016ws}.

\item Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
Society for Research Software}~\cite{GAllen:2016ws}.

\item Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
\emph{Experience Paper: Preliminary analysis of a survey of UK Research
Software Engineers}~\cite{Philippe:2016ws}.

\item Christopher Gwilliams: \emph{Demo: Using industrial engagement to
create and develop research ties within academia}~\cite{Gwilliams:2016ws}.

@@

\noindent The second main area, concerned with characteristics and needs of research software, includes the following five papers:

\begin{itemize}

\item Anshu Dubey\textsuperscript{\textasteriskcentered} and Katherine Riley: \emph{Experience Paper: Software
Engineering and Community Codes Track in ATPESC}~\cite{Dubey1:2016ws}.

\item Neil Chue Hong: \emph{Position Paper: Why do we need to compare
research software, and how should we do it?}~\cite{ChueHong:2016ws}.

\item Anshu Dubey\textsuperscript{\textasteriskcentered} and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
of Software for Scientific Simulation}~\cite{Dubey2:2016ws}.

\item Francisco Queiroz\textsuperscript{\textasteriskcentered} and Rejane Spitz: \emph{Position Paper:
Collaborative Gamification Design for Scientific Software}~\cite{Queiroz:2016ws}.

\item Bruce Childers\textsuperscript{\textasteriskcentered}, Jack Davidson, Wayne Graves, Bernard Rous, and
David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
the Way Digital Libraries will Operate}~\cite{Childers:2016ws}.

@@

\noindent Three contributions focused on the third main area: elucidating lessons learned
from or visions for use cases of scientific software and communities working
with and\slash or on a software package.

\begin{itemize}

\item Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse\textsuperscript{\textasteriskcentered}: \emph{Demo: Composing,
Reproducing, and Sharing Simulations}~\cite{Ganguly:2016ws}.

\item Sameer Shende\textsuperscript{\textasteriskcentered} and Allen Malony: \emph{Demo: Using TAU for
Performance Evaluation of Scientific Software}~\cite{Shende:2016ws}.

\item Janos Sallai\textsuperscript{\textasteriskcentered}, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
\emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
Atom-Typing for Molecular Simulation}~\cite{Sallai:2016ws}.
